### 2010-10-24 Thesis/Chapter10.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Physics-based reconstruction using shell elements}
\label{chap10}
\begin{shortAbstract}
In the previous chapter, we proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach. This simple shell element can efficiently handle both membrane and off-plane forces (bending and twist). The validity of our approach has been demonstrated though comparison with theoretical results. It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation. These results are very encouraging and show the potential of such models. Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels. However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches. In the next section, we propose to study how to approximate the surface of anatomical structures with (curved) shell elements.
\end{shortAbstract}

\section{Techniques of mesh simplification}

\subsection{Introduction: our motivation}

Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe a given geometry with the same precision since they can be curved. Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry. Literature about volumetric mesh generation algorithms is fairly dense. However, there are only a few that are concerned about the generation of meshes over curved surfaces. One of the most widely used techniques for the creation of surface meshes is the plane to surface transformation method \citep{Zienkiewicz71}, mesh is first generated on a two-dimensional domain and then mapped onto the surface. If this method gives reasonably good meshes on smooth surfaces, the results are usually quite poor with more complex curved surfaces. The finite elements may be generated directly on the curved surfaces based on the advancing front technique \citep{Lo85,Lau96}. The main issue with this approach is that an analytical description of the geometry is needed, which is not the case in medical simulation. Another method consists of using an a priori error estimator to build an adaptive mesh generation \citep{Baumann97}. However, the tolerance of this indicator should be chosen depending on the desired accuracy of the finite element solution. Therefore it requires some knowledge about the problem in order to choose an effective tolerance. \cite{Bechet02} start from an existing triangular mesh created with a CAD (computer-aided design) software and refine and smooth the mesh based on element quality and surface curvature. While all those methods allow their authors to get satisfactory meshes over curved surfaces according to their needs, they all make use of flat elements to mesh geometries and do not generate actual shells.

In fact, it is interesting to note that this process of meshing a curved surface is similar to the reconstruction of the surface of objects in computer vision. Indeed, calculating curvature maps of 3D surfaces represented as digitised data (point clouds or triangulated meshes) has been extensively studied. One of the most common approach is to use continuous surface patches \citep{Kolb95,Douros02}. The surface is locally approximated by an analytic representation called surface patch, usually chosen to be an implicit surface. \cite{Tang99} presented an elegant and non noise-sensitive approach that infers sign and direction of principal curvatures directly from the input and the authors used this information for coherent surface and curve extraction from 3D data.

However, we need to realise that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure. In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell. By keeping this in mind, we propose an approach to mesh a curved surface based on the same polynomial shape function \eqref{chap9:eq-deflection} used in our co-rotational shell FEM.

%Simplification strategies may be grouped into two categories: local strategies that iteratively simplify the mesh by the repeated application of some local operator, and global strategies that are applied to the input mesh as a whole \citep{Cignoni97,Talton04}.
%
%\subsection{Local simplification strategies}
%
%Typically, they define some mesh operation $S$ that, when applied to a mesh $M$, acts on a small collection of its elements and produces a new mesh $\overline{M}$  with fewer elements. By repeated application of $S$, a mesh may be simplified arbitrarily. In order to determine the mesh elements to which $S$ should be applied on a given iteration, $S$ may be associated with an error function (or cost function) that measures the amount of error the operation will introduce into the approximation. By computing the error associated with every possible application of $S$ at a particular iteration, the algorithm can apply the one with minimal cost. This type of heuristic is quite reasonable for simplification problems, and in practice these methods work well.
%
%One nice property of local iterative algorithms is that they allow the user to specify the desired attributes of the target approximating mesh with a high degree of precision. For example, the user may allow the algorithm to run until the mesh contains $k$ faces, or until the error at a given vertex exceeds some threshold $\varepsilon$. 
%
%\subsubsection*{Vertex decimation.}
%The decimation algorithm, first proposed by \cite{Schroeder92}, is simple. Multiple passes are made over all vertices in the mesh. During a pass, each vertex is a candidate for removal and, if it meets the specified decimation criteria, the vertex and all triangles that use the vertex are deleted. The resulting hole in the mesh is patched by forming a local triangulation. The vertex removal process repeats, with possible adjustment of the decimation criteria, until some termination condition is met. 
%
%The error associated with a particular decimation typically depends on the classification of the vertex being decimated. For the general case of a manifold vertex $\lbrace i \rbrace$ not near a boundary, \citeauthor{Schroeder92} consider the set of planes formed by the adjacent triangles and compute a single approximating plane based on the area-weighted average of the triangle normals $\textbf{n}_k$, centers $x_k$, and areas $A_k$:
%%
%\begin{equation}
%\textbf{n} = \frac{\sum_k A_k\textbf{n}_k}{\sum_k A_k},  \hspace*{1cm} \widehat{\textbf{n}} = \frac{\textbf{n}}{\vert \textbf{n} \vert}, \hspace*{1cm}  x = \frac{\sum_k A_k\textbf{x}_k}{\sum_k A_k}
%\end{equation}
%
%Then they calculate the distance from vertex $\lbrace i \rbrace$ to this plane by  $\vert \widehat{\textbf{n}} \cdot (v_i - x) \vert$ and define this quantity to be the cost associated with the decimation. This metric ensures that vertices in smooth regions will be decimated before vertices that define sharp features. 
%
%In practice, simplification schemes based on vertex decimation excel at eliminating un-necessary geometry, such as is typically found in meshes generated by the marching cubes algorithm where large planar regions may be subdivided into many redundant triangles. 
%
%\subsubsection*{Edge contraction.}
%Edge contraction, originally proposed by \cite{Hoppe93}, is the most common simplification operation. An edge contraction operates on a single edge $\lbrace v_1,v_2 \rbrace$ and contracts that edge to a single vertex $\lbrace w \rbrace$, updating all edges previously incident on $\lbrace v_1 \rbrace$ and $\lbrace v_2 \rbrace$ to reference $\lbrace w \rbrace$. On a manifold mesh without boundary, each edge contraction will collapse precisely two faces (see \fig{chap10:fig-edgeCollapse}). Edge contractions can alter the topology of a mesh, since repeatedly contracting all the edges around a hole will eventually close it.
%%
%\begin{figure}
%\centering
%\includegraphics[height=3cm]{chapter10/edgeCollapse.pdf}
%\caption[The edge contraction operation]{The edge contraction operation. Image courtesy of \cite{Hoppe93}.}
%\label{chap10:fig-edgeCollapse}
%\end{figure}
%
%For a given contraction $\lbrace v_1,v_2 \rbrace \rightarrow \lbrace w \rbrace$, it is not immediately clear what value should be assigned to $v_h$, i.e. where the resulting vertex should be placed. Obvious choices such as $v_h = v_i$, $v_h = v_j$, or $v_h = (v_i + v_j)/2$ are convenient, but can easily be shown to be non-optimal. Rather than placing $v_h$ arbitrarily, it is better to consider the error function associated with the contraction operation and attempt to minimise its value over the space of possible vertex placements. Once the optimal target position has been computed for every possible contraction, the contraction with the smallest associated error can be selected and applied.
%
%\subsubsection*{Adaptative subdivision.}
%Coplanar or nearly coplanar faces are searched for in the mesh, merged into larger polygons, and then retriangulated into fewer faces than those originally required \citep{Dehaemer91,Hinker93}. Face merging is driven by a co-planarity test. 
%
%The basic adaptive subdivision method as presented by \cite{Schmitt86} is sketched in \fig{chap10:fig-adaptativeSubdivision}. (a) shows several data points produced by the sampling process and sketch (b) adds a trial polygon that approximates the surface represented by the sample points. A bilinear interpolation is performed over the surface of the approximating polygon to locate the point on the surface 'directly below' a sample data point. The distance (or error) between the surface and the data point is then calculated, as shown in (c), and is compared to the user specified tolerance. If any of the errors exceeds the tolerance value then the polygon is divided into four smaller polygons, as shown in (d), and the process is repeated recursively. 
%%
%\begin{figure}
%\centering
%\includegraphics[width=13.5cm]{chapter10/adaptativeSubdivision.pdf}
%\caption[Surface-fitting polygons with adaptive subdivision]{Surface-fitting polygons with adaptive subdivision. Image courtesy of \cite{Schmitt86}.}
%\label{chap10:fig-adaptativeSubdivision}
%\end{figure}
%
%When the approximation to the data is within the specified tolerance, the recursion is terminated and the polygon is saved in a linked list. Upon completion of the subdivision algorithm, the resulting polygons are displayed for the user to judge the results. Generally, the user runs the program repeatedly while varying the maximum tolerance until the results are suitable.
%
%\subsubsection*{Polygon growth.}
%Another method proposed by \cite{Dehaemer91} to systematically generate surface-fitting polygons is to use a polygon growth technique (see \fig{chap10:fig-polygonGrowth}). First a seed polygon is selected from the original set of data, as in (a). Then a neighbouring polygon is selected and the two are combined into a larger trial polygon ((b) and (c)). If the trial polygon passes the accuracy metric, 10 additional neighbors are selected in an attempt at further growth, as in (d) through (g). When all attempts at further growth fail, the polygon is added to the list of polygons describing the reduced object, as shown in (h). A new seed polygon is selected and the process repeated. When there are no more polygons from which to choose a new seed, the process is terminated.
%%
%\begin{figure}
%\centering
%\includegraphics[width=12cm]{chapter10/polygonGrowth.pdf}
%\caption[Polygon growth]{Polygon growth. Image courtesy of \cite{Dehaemer91}.}
%\label{chap10:fig-polygonGrowth}
%\end{figure}
%
%This method of fitting polygons to the sampled data is not as useful as the adaptive subdivision methods, for several reasons. First, it is an essentially brute force method of surface fitting and very repetitive in nature. Many trial polygons are created, only to be 'thrown out' because they exceed the error tolerance. This causes the procedure to execute quite slowly (orders of magnitude more slowly) than the adaptive subdivision method. Second, the data is not significantly reduced. Many of the generated polygons are of complicated shapes and use a large number of vertices. Thus many of the original sample data points are retained.
%
%
%\subsubsection*{Appearence preserving simplification.}
%A completely different approach to determining the error associated with a given simplification operation was described by \cite{Cohen98}. Their metric, which is appearance-based, measures the amount of deviation caused by the operation in the screen-space (pixelised) representation of the mesh. Their algorithm functions by decoupling surface position from color and curvature information and storing the latter two quantities in texture and normal maps. A traditional geometric simplification algorithm can then be employed to filter the surface position, while a hardware-based approach is used to filter the color and normal information, resulting in simplified representations that are nearly visually indistinguishable from the original. However, the algorithm requires a parameterisation of the mesh in order to function.
%
%
%\subsection{Global simplification strategies}
%
%\subsubsection*{Vertex clustering.}
%The method of vertex clustering was originally proposed by \cite{Rossignac93} to handle meshes of arbitrary topological structure. In their algorithm, each vertex in the input mesh is assigned a weight based on its perceptual importance: vertices adjacent to triangles with large faces and those in areas of high curvature are weighted more heavily than vertices in smooth regions adjacent to smaller triangles. Next, a bounding box is placed around the mesh and subdivided into a three-dimensional grid. Finally, all the vertices in a given grid cell are clustered to the position of the vertex with maximum weight. Degenerate geometry may then be removed, although the authors proposed a novel visualisation of certain degenerate faces and edges in an effort to enhance the recognizability of their simplified meshes. The algorithm is extremely efficient and simple to implement, and the level of simplification may be controlled (although with some difficulty) by choosing the resolution of the overlaid grid. However, vertex clustering can drastically alter the topology of the input mesh in an unpredictable manner, and in practice does not produce very faithful geometric approximations at low levels of detail.
%
%\subsubsection*{Re-tiling.}
%The re-tiling method described by \cite{Turk92} begins with a polygonal surface and creates a triangulation of this surface with a user-specified number of vertices. The points are placed uniformly over a given polygonal surface by distributing points at random over the surface and then having each point repel all of its neighbors. An alternative to placing the points randomly is using the variation in surface curvature. The features of the surface would be more accurately reflected in a re-tiling by increasing the density of vertices in regions of high curvature.
%
%The key notion in the re-tiling procedure is the creation of an intermediate model called the mutual tessellation of a surface that contains both the vertices from the original model and the new points that are to become vertices in the re-tiled surface. The new model is then created by removing each original vertex and locally re-triangulating the surface in a way that matches the local connectedness of the initial surface (see \fig{chap10:fig-retiling}).
%
%If this method produces nice meshes, an issue is finding measures of how closely matched a given re-tiling is to the original model. 
%%
%\begin{figure}
%\centering
%\includegraphics[width=12cm]{chapter10/retiling.pdf}
%\caption[Retiling technique]{Upper left: original surface. Upper right: candidate vertices after point-repulsion. Lower left: mutual tesselation. Lower right: final tesselation. Image courtesy of \cite{Turk92}.}
%\label{chap10:fig-retiling}
%\end{figure}

\section{Our method}

	\subsection{A simple algorithm}

In the following, we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the object we want to simulate (via a Marching Cube algorithm for instance, \cite{Lorensen87}). We propose a technique that can simplify the high resolution mesh of the object to be simulated and approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation. Thus, the first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in \citeauthor{Meshlab}. MeshLab is an open source program for the processing and editing of unstructured 3D triangular meshes. The algorithm tries as much as possible to preserve mesh boundaries and generates high quality triangular elements. We then apply a heuristic method derived from the work of \cite{Saupin07} with tetrahedral meshes based on simple geometrical rules. For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycentre of the three centres of mass of those triangles. This technique locally smoothes the surface of the mesh while converging towards the desired high resolution mesh.

At each iteration of this algorithm, we want to measure the error between the curved surface of shells and the target. Indeed, we need to ensure that the distance between the surface of our shell-based mesh and the targeted high resolution mesh will be minimal. An efficient technique for measuring the error between two surfaces is the Hausdorff distance \citep{Klein96}. As a reminder the Hausdorff distance between two meshes is the maximum between the two so-called one-sided Hausdorff distances:
\begin{equation}
d_{\mathrm{H}}(X,Y) = \max \left\{ \sup_{x \in X} \inf_{y \in Y} d(x,y), \sup_{y \in Y} \inf_{x \in X} d(x,y) \right\},
\end{equation}
where $d()$ is the Euclidian distance between two points. The same technique of subdivision used for rendering allows us to sample the actual surface described by the shells to compute the Hausdorff distance with the targeted high resolution mesh.

By using the Hausdorff distance between two iterations, the process may be stopped when the required precision has been reached. A simple example is shown \fig{chap10:fig-cylinder} to illustrate the method. 
\begin{figure}[ht]
\centering
\includegraphics[width=13.5cm]{chapter10/exampleCylinder.pdf}
\caption {The target (a) is a high resolution cylinder mesh of 16,384 triangles and we start from a very coarse mesh (12 triangles), rendered with flat triangles here (b). In (c) the coarse mesh is rendered with shells and a one-sided Hausdorff distance colour map is applied to show the initial error with the high resolution mesh. (d) One-sided Hausdorff distance colour map after one iteration of our algorithm (48 shells).}
\label{chap10:fig-cylinder}
\end{figure}

	\subsection{Results on complex anatomical structures}

\subsubsection*{Meshing of anatomical structures.}
This approach has been applied to approximate more complex anatomical geometries with curved shell elements. Two examples are given with the Glisson's capsule, which is the membrane surrounding the liver (\fig{chap10:fig-liver}) and an aneurysm (\fig{chap10:fig-aneurysm}). In each case, the error is computed through a Hausdorff distance and expressed as a percentage of the diagonal of the object's bounding box. 
%
\begin{figure}[ht]
\centering
\includegraphics[width=13.5cm]{chapter10/resultsLiver.pdf}
\caption {(a) the targeteted high resolution Glisson's capsule mesh ($8\,000$ triangles). (b) the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh ($1\,200$ shells).} 
\label{chap10:fig-liver}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=13.5cm]{chapter10/resultsAneurysm.pdf}
\caption {(a) the targeteted high resolution aneurysm mesh ($28\,368$ triangles). (b): the one-sided Hausdorff distance error map on a mesh of 772 shells generated with our method.}
\label{chap10:fig-aneurysm}
\end{figure}

\subsubsection*{Computation times.}
We perform several tests on the aneurysm model at different resolutions to measure computation times. The shells are resisting to a uniform pressure load and solved using a Conjugate Gradient (CG) iterative solver. The computation times are reported in \fig{chap10:fig-computation}. Implicit integration allows for large time steps ($40\,ms$) and the computation is real-time for 800 shell elements and a reasonable error criterion ($5\,\%$). When the computation time must be bounded (critical real-time applications), one can fix the number of CG iterations to, for instance, $100$ and remains real-time for $1\,000$ shell elements. However, in that case the accuracy of the results is not checked.
%
\begin{figure}[ht]
\centering
\includegraphics[width=13.5cm]{chapter10/computation_times.pdf}
\caption {Computation time on meshes of 200, 400, 600, 800 and $ 1\,000 $ elements.}.
\label{chap10:fig-computation}
\end{figure}

\subsection{Coupling between tetrahedra and shells for advanced modelling}
Structures in human body can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.). However, knowing how to model the two kind of structures is not sufficient to reach a high degree of accuracy, real life situations are more complex. As an example, the external surface of the liver is covered by a layer of cells called Glisson's capsule. Its interaction with the liver plays an important role into the overall structure's mechanical behaviour. Therefore considering the interaction between solid and hollow objects is as crucial as modelling the two structures separately.

An example of medical procedure to illustrate this point even further is angioplasty. Angioplasty is the technique of mechanically widening a narrowed or obstructed blood vessel, typically as a result of atherosclerosis. An empty and collapsed balloon on a guide wire is passed into the narrowed locations and then inflated to a fixed size. The balloon crushes the fatty deposits, so opening up the blood vessel to improved flow. As a proof of concept, we tried to simulate an angioplasty. Our preliminarty results are presented on \fig{chap10:fig-stent}. The blood vessel is modelled using the shell FEM formulation described in this paper and the fatty deposits are simulated with a tetrahedral FEM method and are fixed to the interior wall of the blood vessel. When the balloon inflates it crushes the deposits and they then apply a pressure onto the curved surfaces of shells modelling the interior wall. The forces are then distributed onto the mechanical nodes of the blood vessel mesh as detailed in section~\ref{chap9:interactions}, which widens the blood vessel as expected. 
%
\begin{figure}[ht]
\centering
\includegraphics[width=12cm]{chapter10/stenose_final.pdf}
\caption {Simulation of an angioplasty procedure. (a, c): A collapsed stent is inserted into the blood vessel simulated with our shell FEM formulation. The fatty deposits (in yellow) are modelled with tetrahedral FEM. (b, d): The stent is crushing the fatty deposits which creates a pressure onto the interior wall and widens the blood vessel.}
\label{chap10:fig-stent}
\end{figure}

@@

Although we were able to mesh complex anatomical structures with a fairly small number of shell elements, our approach could be further improved. In our current implementation, if the error measured with the Hausdorff distance is greater than a threshold (which is also difficult to choose), all triangular elements are subdivided into 4 new triangles and each of the new vertices is moved towards the targeted geometry based on our heuristic algorithm. Obviously, this is not optimal as the error between the two meshes may very well be local and therefore the use of a global mesh refinement is not very suitable. The second weakness is that we also do not check the quality of the elements during our mesh process. If some elements are stretched from the heavy decimation, these elements are not corrected during the process of re-meshing. Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements. One solution for the latter issue would be to enhance our algorithm with a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object). This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall. Yet, meshing a surface with curved elements is a very complex problem and still an active area of research. In contrast to all previous geometrical techniques, we suggested a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.

\section{Conclusion}

Over the last two chapters, we proposed a complete framework for real-time modelling of thin anatomical structures. The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells. We also show how contacts and interactions with the curved surfaces of shells can be handled using the same function. In summary, these polynomial shape functions are used in three different ways in our computational model:
%
\begin{description}
\item[(a)] to approximate complex geometrical shapes
\item[(b)] to compute internal forces via a co-rotational shell FEM
\item[(c)] to compute contact forces onto a curved triangle
\end{description}

We achieved our objectives to propose a versatile solution which can simulate, in real-time, thin objects with various shapes and material properties with good accuracy. The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects (like an intra-ocular implant). We also presented preliminary results on the coupling between solid (tetrahedra) and thin objects (shells) for the advanced modelling of anatomical structures via the simulation of an angioplasty procedure. This work on modelling thin anatomical structures was the object of two main publications \citep{Comas2010a,Comas2010b}.

>>>>

\chapter{Conclusion and perspectives}
\label{chap10}

\section{Summary of contributions}

The objective of this PhD was to propose soft tissue models for computing the deformation of anatomical structures in real-time. Because the shape and the internal nature of these structures greatly differ, a unique model cannot deal with the variousness of the mechanical behaviours. Hence, we identified two major types of structures in human body, they can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).

Most of previous works focused on volumetric models that are able to capture the behaviour of solid organs. Yet, many of these models had to compromise on accuracy to achieve the real-time constraint demanded by medical simulation. Even though the best mathematical framework for predicting the deformation of a continuous structure is provided by continuum mechanics, solving these equations is a computationally very demanding task and simpler models are often used in practice. And even when the finite element method is employed to solve the equations of continuum mechanics, linear strain measure and simple constitutive laws are common and therefore does not constitute an appropriate modelling of the complex mechanical behaviours of solid anatomical structures.

In contrast, much less research has been carried out for the modelling of hollow structures. In fact in the field of medical simulation, very few models have been proposed for simulating, in real-time, the deformation of thin anatomical structures whose volume is negligible compared to their surface area. Nevertheless, human body features many hollow structures and we feel that the need for modelling hollow organs in medical simulation is equally important as with solid organs.

\bigskip

Consequently, we strived to propose solutions to these two known problems in the field of medical simulation.

\subsubsection*{Solid structure modelling.}
We implemented the total Lagrangian explicit dynamics (TLED) algorithm on GPU using CUDA. This finite element formulation is fully non-linear: it makes use of a non-linear strain measure (which allows large deformations) and embodies a non-linear constitutive law between strain and stress as characterised by \cite{Fung93} for soft-tissues. While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, our efficient re-implementation in CUDA led to a threefold increase in performance. Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up to $56.3\,$x. My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity. To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure. This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publicly with the upcoming 1.0 release of SOFA. 
%By providing a fast and accurate non-linear FEM with various constitutive laws for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.

\subsubsection*{Hollow structure modelling.}
We proposed a framework for real-time modelling of thin anatomical structures. We first designed a triangular shell element by combining a two-dimensional in-plane membrane energy, with an off-plane energy for describing bending and twist. We improved and extended a plate model first introduced by \cite{Przemieniecki85} to a co-rotational formulation to allow large displacements (but small deformations). Once combined with an in-plane membrane formulation we obtained an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies. To the best of our knowledge, this co-rotational shell FEM is the first description of a shell finite element method in the field of medical simulation. However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches. Indeed, while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe a given geometry with the same precision since they can be curved. Therefore, we suggested a simple method to mesh a surface with curved shell elements. The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells. We also showed how contacts and interactions with the curved surfaces of shells can be detected and handled using the same function.

@@

	\subsection{Solid structures modelling}

The TLED algorithm offers an accurate modelling for soft tissues and its GPU implementation using CUDA is very efficient. Hence, real-time computations can be achieved without any trade-off on the accuracy of the deformation. One of the features allowing such an efficiency is the use of an explicit time integration scheme. If this permits significant computational advantage, the use of an explicit integration scheme imposes that the time step of the simulation remains below a critical time step. This critical time step depends on two kinds of criteria: the size of the elements and the material's properties (Young's modulus and Poisson's ratio). Therefore, the meshing of the anatomical structures to be simulated must be accomplished with great care to insure well-shaped elements. Even more restricting, the use of the TLED formulation is not suitable for modelling stiff objects in real-time. Indeed, increasing the Young's modulus translates into a lower critical time step, which necessitates an increase of the simulation's frequency to enforce real-time. In some applications, the value of the Poisson's ratio may be relaxed from 0.49 to 0.45 for instance. However, if it allows to double the critical time step, this relaxation introduces additional inaccuracies.

One of the most common critics we received for this work concerned the poor ability of the TLED to handle stiff contacts. Indeed, interactions may impact the critical time step resulting from the use of an explicit time integration scheme and thus influence the overall stability of the simulation. The work carried out during my PhD was entirely concerned with the numerical solution provided by the TLED formulation. Yet, while this is beyond the scope of the present work, a range of contact formulations are available in commercial finite element packages. According to their documentation, the impact on the critical time step should be minimal. However, for a demonstration at an IT exhibition (CeBIT Australia), we tried to setup an interaction between a virtual tool driven by a haptic device and a cube with brain-like properties. If we managed to produce an interesting demo for the exhibition, the system appeared fairly unstable. But even if a stable contact modelling technique could be formulated, the actual limiting factor is the collision detection process. Because the time step must be very small when using an explicit time integration scheme (often around one millisecond or less), the collision detection algorithm must also be run at the same frequency. The problem is that no collision detection algorithm is fast enough to execute at $ 1\,000\,$Hz or more with objects of reasonable sizes. One notable exception is the interaction with a fixed rigid surface where the collision detection is rapid.

Consequently, this list of pros and cons limits the TLED's range of applications. If its usage in very interactive environments is not suitable, the TLED finite element algorithm is great at computing the deformation of soft tissues when the boundary conditions are known and controlled. The TLED is also a great algorithm for non-rigid registration problems found in image-guided therapy \citep{Modat10}. Indeed, intra-operative image registration must be fast enough that the work flow of the surgical procedure is not interrupted. Hence, provided that the application is appropriate, the TLED algorithm is an efficient and accurate algorithm to model the deformation of solid soft anatomical organs.

	\subsection{Hollow structures modelling}

We achieved our objectives to propose a versatile solution which can simulate, in real-time, thin objects with various shapes and material properties with good accuracy. The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects. Nevertheless, the use of a linear strain measure limits the range of application to small deformations (and large displacements). Moreover, the constitutive law was chosen to be the simple (linear) Hookean law, which it is known not to describe accurately the relationship between strain and stress in soft tissues. However, those potential changes are substantial and their implementations are not straightforward. In addition, it is not clear from which method the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true mathematical shell theory (where membrane and bending energies are not separated). Our shell FEM could also be improved by implementing the formulation on GPU. Although an optimal parallel implementation is always challenging, our method is highly parallelisable and there is no major obstacle to an efficient GPU implementation.

Regarding our geometric surface reconstruction based on shells, we were able to mesh complex anatomical structures with a fairly small number of shell elements. However, the main drawback of our method is the lack of any assessment of the quality of the elements created. Indeed, if some elements are stretched from the heavy decimation, these elements are not corrected during the process of re-meshing. Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements. Yet, meshing a surface with curved elements is a very challenging problem and still an active area of research. In contrast to all geometrical techniques described in the literature, we presented a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.

	\subsection{Use of GPU acceleration}

One of the strongest constraints in medical simulation is the demand for real-time computations. This constraint is what makes all tasks in a simulator so challenging (collision detection, contact and soft tissue modelling, haptic feedback, photo-realistic rendering etc.). In fact, this is at the origin of the multitude of simplifications proposed in simulation for modelling the biomechanics of anatomical structures. The lack of accuracy induced by these simplifications is of great concern to us. In this PhD, a part ot the solution was to accelerate the computations by using the GPU. While we have showed that this approach can work very well, we need to keep in mind that GPU acceleration is not the ultimate solution. Firstly, a standard is yet to be imposed in GPGPU. While OpenCL is openly targeted as a multi-vendor and multi-platform API, its usage remains very scarce nevertheless. In fact, the first experiments carried out within the SOFA framework revealed that OpenCL is not as optimised as CUDA for instance (up to twice as slow). Secondly, GPU programming is not simple to master. Although we have seen many improvements in flexibility of programming over the last 3 or 4 years, a strong understanding of the underlying GPU hardware is still mandatory to create an optimal code. As an example, a GPU offers different types of memory, all with distinct pros and cons (strict access patterns, latency, limited size etc.), and choosing the best type of memory for each variable may be rather challenging to say the least. But most importantly, GPU acceleration is only interesting if the code is highly parallelisable. The difficulty is to devise a model that combines (1) an excellent capability at predicting the deformation of anatomical structures and (2) great parallel properties. Ideally, the parallelisation of the algorithm must be thought as part of the design process rather than seen as two unconnected steps. Even when the code is greatly parallelisable, one could face limited GPU hardware possibilities (as with concurrent writes for instance). GPU acceleration is all but a straightforward process. One must often accomplish a considerable amount of work to achieve a substantial gain. 
% However, a decent GPU implementation may allow improvements of up to one or two orders of magnitude. For instance, our CUDA implementation of the TLED algorithm was about $ 56 \times $ faster than an equivalent CPU implementation on 2008 hardware.

\section{Perspectives}

Earlier this year, NVIDIA released a new GPU architecture called \emph{Fermi}. Fermi features several major innovations for GPGPU computing like cache (faster memory access in some cases), unified address space (easier to run C and C++ programs), debugging support, faster atomic instructions, compliance with the newly revised IEEE 754 floating point standard, three times the amount of shared memory, etc. In theory, this series of improvement should boost the performance of the second and third TLED implementations (see section \ref{chap6:secondTLED}) and we should investigate and compare the efficiency of the three designs on this new hardware.

In a short term perspective, our co-rotational shell FEM algorithm could also be implemented on GPU. We already assessed in section \ref{chap8:GPU} that there was no major obstacle which would prevent a parallel implementation. Nevertheless, an optimal GPU implementation will demand careful design to take full advantage of the hardware.

Although the meshing of complex surfaces with curved shell elements was not the heart of this PhD, our naïve approach could be enhanced by focusing on element quality. One simple solution would be to add a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object). This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall. While a more mathematical method (a derivative of advancing front techniques for instance) could probably be formulated, the problem is complex and could certainly account for another PhD.

Finally, all the contributions of this PhD could contribute to enhance the complexity of anatomical structure modelling through combining solid and hollow structures. While we presented our findings on the coupling between solid (tetrahedra) and thin objects (shells) via the simulation of an angioplasty procedure, these results are still preliminary and much work remains to be done. In particular, the validity of the interaction and the accuracy of the overall deformation of the blood vessel were not quantitatively evaluated. Another interesting example of coupling would be the complete model of the liver. The relationship between strain and stress within the parenchyma (functional part of the liver) is known to be non-linear and anisotropic. Moreover, both the vascular network within the tissue and the Glisson's capsule have a strong influence on the overall mechanical behaviour of the liver. The contributions introduced in this PhD thesis may offer a solution to this challenging problem. While the parenchyma could be simulated with an algorithm such as the TLED (materially non-linear and able to account for the anisotropy), the blood vessels and the Glisson's capsule could be described using our co-rotational shell FEM.

As a conclusion, the TLED algorithm and our co-rotational shell FEM are respectively the first non-linear FEM for solid structures and the first model featuring bending for hollow structures to be implemented into the open source framework SOFA. Although they are based on continuum mechanics, they may both be employed to simulate the deformation of entire organs in real-time. By providing efficient FEM algorithms for all types of anatomical structures to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.
